Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-11 joint_inf.tex
add goldman and allman/chai

An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the states of ancestral sequences.
%page 2 section 2.1 at the top of the second column first paragraph second sentence
Although the software described in \citet{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

@@

Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \citep{Murphy2000-ry}.
Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \citep{Geman1982}.
A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \citep{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \citep{LeCam1990}.
Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
We have furthered progress in understanding the limitations of this joint optimization procedure.

@@

>>>>

An estimator is said to be statistically consistent if it converges to the generating model with probability 1 in the large-data limit; existing consistency proofs for maximum likelihood phylogenetics \citep{Allman2008-wd,Chai2011-ff,RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from problems where the number of parameters increases with the amount of data \citep{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the states of ancestral sequences.
%page 2 section 2.1 at the top of the second column first paragraph second sentence
Although the software described in \citet{Sagulenko2017-jo} fits on a user-supplied topology and the authors explicitly warn that the approximation is for the case where ``branch lengths are short and only a minority of sites change on a given branch,'' their work motivates understanding the general properties of such joint inference.
In particular, one would like to know when this approximate technique breaks down for both topology and branch length inference, even when sequence data is ``perfect,'' i.e., is generated without sampling error according to the exact model used for inference.

@@

Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \citep{Murphy2000-ry}.
Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., the method of sieves \citep{Geman1982}.
A potential solution in this case using the method of sieves could be to project the column-wise ancestral state patterns into a lower-dimensional space, allowing the degrees of freedom in the ancestral state columns to grow with $\nCols$, albeit more slowly than $O(\nCols)$.
Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \citep{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \citep{LeCam1990}.
Consistency proofs of standard maximum likelihood estimates of phylogeny \eqref{eq:marginal_likelihood} are recent \citep{Allman2004-wd,Chai2011-ff,RoyChoudhury2015-ta}, and no results have been obtained for profile likelihood.
We have furthered progress in understanding the limitations of this joint optimization procedure.

@@

The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
In it, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
We show cases where the estimates of the topology will be inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths will also be inconsistent.
Finally, just prior to his conclusion, he discusses when parsimony will give the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood, rendering the question well-posed.
Finally, we are able to provide much more detail on how large branch lengths must be for general joint inference---estimating topology and branch lengths together---to fail to be consistent.
